---
title: "PCA with stars objects"
author: "Joseph Holler"
---

```{r, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, collapse = TRUE, dev = "png")
require(dplyr)
require(stars)
require(here)
```

This vignette is based on sections of Edzer Pebesma's Statistical modelling with stars objects: <https://r-spatial.github.io/stars/articles/stars7.html>

### stars objects as data.frames

The analogy of stars objects to `data.frame` is this:

* each attribute (array) becomes a single column
* dimensions become added (index) columns

To see how this works with the 6-band example dataset, consider this:

```{r}
library(stars)
l7 <- read_stars(here("L7.tif"))
l7
as.data.frame(l7) %>% head()
```

We see that we get **one** single variable with the object (array) name, and
added columns with the dimension values (x, y, band). In a typical case, we
would like to have the six bands distributed over six variables, and have
a single observation (row) for each x/y pair.
For this, we _could_ use e.g. `utils::unstack` or `dplyr::pivot_wider` on this data.frame, but
a more efficient way is to use the dedicated `split` method for `stars` objects,
which resolves a dimension and splits it over attributes, one for each dimension value:

```{r}
l7 %>% split("band") %>%
  as.data.frame() %>% 
  head()
```

The reason that `split` is more efficient than the mentioned alternatives is that (i) `split` does not have to match records based on dimensions (x/y), and (ii) it works for out-of-memory (stars_proxy) arrays, in the chunked process/write loop of `write_stars()`.

### Principal components

First, let's visualize six bands of data included in a Landsat 7 scene.

```{r}
plot(l7, breaks = "equal", join_zlim = FALSE)
```
As you can see, many of the bands contain similar information. 

We can visualize information from at most three bands at a time with red (band 3), green (band 2), and blue (band 1) color.
For example, here is a true color visualization with red color = red energy, green = green energy, and blue = blue energy.

```{r}
plot(l7, rgb = c(3,2,1) )
```

Or, here is a false color visualization with red color = infrared energy, green = red energy, and blue = green energy.
In these visualizations, vegetation appears red and urban areas appear blue-green.

```{r}
plot(l7, rgb = c(4,3,2) )
```

What if we could create **one** band that contains the most possible information from all six?

And then a **second** band containing the most possible remaining information?

and so on...

We probably wouldn't even need all six bands to convey most of the information... 

Let's calculate the principal components with **principal component analysis (PCA)**.

```{r}
tif = system.file("tif/L7_ETMs.tif", package = "stars")
r = split(read_stars(tif))
pc = prcomp(as.data.frame(r)[,-(1:2)]) # based on all data
out = predict(r, pc)
write_stars(merge(out), "pca.tif")
```

Display the variance explained by each component

```{r}
summary(pc)
```

The first component explains 70% of variance in the six satellite image bands.  
The second component explains 25% of the variance, for a cumulative 95% of all variance.  
By the time we add a third component, we have already explained 99.7% of all of the variance from six bands of satellite data.

Let's display the loadings for each principal component

```{r}
pc$rotation
```

How can we interpret this?
The first component is dominated by information from Bands 5 and 6 (short wave infrared)
The second component is dominated by a lack of visual light (bands 1 2 and 3) and presence of near infrared (band 4)

```{r}
plot(merge(out), breaks = "equal", join_zlim = FALSE)
```

First of all, the band with the most contrast is PC1, in which water/land is very clear.  
In PC2 and PC3, we can see more subtle features, including clouds/smoke, beaches, and vegetation / bare or built up areas.
PC3 is starting to emphasize less important elements, including shadows in the topography and differences in the water surface.
PC4 and above have very little contrast and increasing amounts of noise.

Let's rescale the first three principal components for visualization in a single RGB image.
I have arbitrarily inverted the second and third bands to improve the visualization.
That is ok, because PCA loadings are arbitrarily negative or positive in direction.
It's the magnitude of rotation that matters, not the direction (+/-).

```{r}
outmin <- min(out$PC1)
outmax <- max(out$PC1)
out$PC1 <- (out$PC1 - outmin) / (outmax - outmin) * 255
outmin <- min(out$PC2)
outmax <- max(out$PC2)
out$PC2 <- 255- (out$PC2 - outmin) / (outmax - outmin) * 255
outmin <- min(out$PC3)
outmax <- max(out$PC3)
out$PC3 <- 255- (out$PC3 - outmin) / (outmax - outmin) * 255
```

What does our false color composite of the first three components look like?

```{r}
plot(merge(out[1:3]), interpolate = TRUE, breaks = "equal", rgb = c(1,2,3), maxColorValue = 255)
```




